---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(dplyr)
library(smallfactor)
```

```{r eval = FALSE, include = FALSE}
pc <-covr::package_coverage()
percent <- round(covr::percent_coverage(pc), 1)
usethis::use_badge(
  "",
  href = 'https://img.shields.io/badge/testcoverage-100percent-blue.svg',
  src  = 'https://img.shields.io/badge/testcoverage-100percent-blue.svg'
)
# ![](https://img.shields.io/badge/testcoverage-100percent-blue.svg)
```





# smallfactor - store factors in bytes and bits rather than an `integer`

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/smallfactor/workflows/R-CMD-check/badge.svg)](https://github.com/coolbutuseless/smallfactor/actions)
<!-- badges: end -->

`smallfactor` is an experiment to see what trade-offs there might be for storing
a factor in bytes or bits instead of an integer.

An R integer is 4-bytes, meaning it could hold a factor with two billion levels,
which seems like overkill.   

`bytefactor` targets factors with up to 256 levels.

`bitfactor` will choose the fewest bits to store a factor for factors with 2 up to
32768 (2^15) levels.


## What's in the box

* `bytefactor()` store factor values in raw bytes rather than an integer.
* `bitfactor()` store factor values as bits within integers - with multiple factors
  stored in a single integer value.
  E.g. a factor with 4 levels could be stored as just 2 bits of information, 
  meaning 16 values could be stored in a single R integer (16 * 2 bits = 32 bits = 4 bytes)

## Limitations

It seems possible that you could write more methods to make these smaller factors 
behave similar to regular R `factor` objects, but this package is not attempting
this (yet).

Internally, there's currently a lot of transferring back-and-forth between 
these small factors and the standard R factor in order to make use of 
the printing and subsetting capabilities of the R factor implementation.  Much
of this back-and-forth could be avoided if effort was expended to do so.

Note: `bitfactor` uses only 31 bits of a 32-bit integer in order to avoid 
issues around `NA_integer_` representation.  This means, for example, that an 
integer can only hold 15 x 2-bit values.  In practice the user is never
expected to notice this or care about it.

## Installation

You can install from [GitHub](https://github.com/coolbutuseless/smallfactor) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/smallfactor')
```

## `bytefactor`


```{r}
small <- bytefactor(c('a', 'b', 'c', 'a', 'd'))
small

small[1:3]
```



## `bitfactor`

`bitfactor()` will choose an appropriate number of bits to store the given 
number of levels.

In the following example, there are 4 levels, so `bitfactor()` chooses to 
store each value in the factor in 2 bits.


```{r}
tiny <- bitfactor(c('a', 'b', 'c', 'a', 'd'))
tiny

tiny[1:3]
```

In this next example, there are 100 levels in the factor, so 7 bits are 
needed to fully store all the levels

```{r}
tiny <- bitfactor(sample(100, 10), levels=1:100)
tiny

tiny[4:6]
```




## Storing some DNA in `bytefactor` and `bitfactor` objects.

|            | character | factor | bytefactor | bitfactor |
|------------|-----------|--------|------------|-----------|
| bits/value | 64        | 32     | 8          | 2         |
| total size | 8 MB      | 4 MB   | 1 MB       | 270 kB    |
| size reduction |       |  2x    | 8x         | 30x       |

```{r example}
library(smallfactor)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Generate some random DNA
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
character_vector_dna <- sample(c('A', 'T', 'G', 'C'), 1e6, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a `factor` and a `smallfactor` using the same basic syntax
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
integer_factor <- factor    (character_vector_dna, levels = c('A', 'T', 'G', 'C'))
byte_factor    <- bytefactor(character_vector_dna, levels = c('A', 'T', 'G', 'C'))
bit_factor     <- bitfactor (character_vector_dna, levels = c('A', 'T', 'G', 'C'))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# `smallfactor` is approx 1/4 the size of the regular factor
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lobstr::obj_size(character_vector_dna)
lobstr::obj_size(integer_factor)
lobstr::obj_size(byte_factor)
lobstr::obj_size(bit_factor)
```

## Similar projects

* [{lofi}](https://github.com/coolbutuseless/lofi)


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
