---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(dplyr)
library(smallfactor)
```

# smallfactor - store factors in `raw` rather than `integer`

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`smallfactor` is an experiment to see what tradeoffs there might be for storing
a factor in a raw byte instead of an integer.

Since an R `raw` value is only a single byte, but an `integer` is 4 bytes, 
there are space saving for large factors if stored in `raw` vector - 
as long as the number of levels is less than 256.

## What's in the box

* `smallfactor()` - similar to `factor()`, but creates an object backed by 
  raw bytes rather than an integer.


## Limitations

This package includes creating an object of the `smallfactor` class, 
and includes a few expected methods for common usage e.g. `as.character.smallfactor()`, 
`as.data.frame.smallfactor()`.

It seems possible that you could write more methods to make `smallfactor` behaviour
almost identical to regular `factor` behaviour, but this package is not attempting
this (yet).

## Installation

You can install from [GitHub](https://github.com/coolbutuseless/smallfactor) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/smallfactor')
```

## Example simple usage


```{r}
small <- smallfactor(c('a', 'b', 'c', 'a', 'd'))
small

small[1:3]

df <- data.frame(value = small, y = 1:5)
df


library(dplyr)
df %>%
  filter(value < 3)
```



## Example - storing some DNA in 1/4 the space of a regular factor


```{r example}
library(smallfactor)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Generate some random DNA
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
dna <- sample(c('A', 'T', 'G', 'C'), 1e6, replace = TRUE)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create a `factor` and a `smallfactor` using the same basic syntax
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
small   <- smallfactor(dna, levels = c('A', 'T', 'G', 'C'))
regular <- factor     (dna, levels = c('A', 'T', 'G', 'C'))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# `smallfactor` is approx 1/4 the size of the regular factor
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
lobstr::obj_size(dna)
lobstr::obj_size(regular)
lobstr::obj_size(small)
```


## Acknowledgements

* R Core for developing and maintaining the language.
* CRAN maintainers, for patiently shepherding packages onto CRAN and maintaining
  the repository
